SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-12-07 papers/B1555_EdgeOn/B1555_letter.tex
Updated paper, with temporary version of new figure

A common feature of dark-matter only simulations of structure formation is the presence of thousands of sub-haloes that are associated with larger mass haloes \citep[e.g.][]{Springel08}. However, observations of the Local Group show many fewer satellite galaxies than are predicted by the simulations, even when taking into account the latest discoveries of new dwarf galaxies \citep{DES15,Kop15} and completeness corrections arising from the limited sky coverage imposed by the Galactic Plane.  
This is the well-known {\it missing satellite problem} \citep{Klypin1999,Moore1999,S07}. In order to understand this discrepancy, it is imperative to build up a large sample of satellites/subhaloes in galaxies outside the Local Group. However, this process is challenging due to the optical/infrared faintness of the satellite galaxies.  Furthermore, lower-mass satellites may not be able to retain the gas needed for ongoing star formation \citep[e.g.][]{P11}, rendering them effectively dark.

@@

B1555+375 was observed as part of SHARP using the NIRC2 camera on the W. M. Keck-II Telescope on 2012 May 16 (PI: Fassnacht).  The AO system was used, with the corrections derived from the laser guide star and a $R=14.4$-mag tip-tilt star that was located $\sim45$~arcsec from the lens.  The narrow camera mode was used, giving a field of view of $\sim10$~arcsec on a side and a pixel scale of 10~mas. Six dithered 300~s exposures were obtained in the $K^{\prime}$-band filter.  The data were reduced with the standard SHARP pipeline, which is a python-based package that is a refinement of the process described by \citet{Auger_EELS1}.  A cutout of the final reduced image is shown in Fig.~\ref{fig:multiband} and also in Fig.~\ref{fig:merlin}, where the contours from 5-GHz Multi-Element Radio Link Interferometer Network (MERLIN) imaging by \citet{Marlow99} and 1.66-GHz VLBA imaging (see below) are overlaid.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, as can be seen by comparing the radio contours to the $K^\prime$-band emission  (Fig.~\ref{fig:merlin}), and the faint, but clearly visible emission from {\bf the lensing galaxy that appears to be an edge-on disc}, with a position angle of $\sim 10$~deg and an {\bf axis ratio $q \sim 0.2$}.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission that is seen from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

B1555+375 was observed with the 10 telescopes of the VLBA at a central observing frequency of 1.66 GHz on 2000 March 21 (BN0009; PI: Norbury). The data were recorded at 128 Mbits~s$^{-1}$ and then correlated to produce two spectral windows with 8 MHz bandwidth, 16 channels and both circular polarizations (RR and LL). The observations were phase referenced using J1544+398 every 3 min over the total observing time of 3 h. The data were reduced in the standard manner using the {\sc vlbautlis} analysis pipeline that is part of the Astronomical Image Processing Software (AIPS). Imaging was done with the {\sc clean} algorithm in AIPS using natural weighting of the visibilities, and restored using an elliptical beam of size $9.7 \times 6.9$~mas at a position angle of $-7.6^\circ$ east of north. The final map (also shown in Fig.~\ref{fig:merlin}) has an rms of 78~$\mu$Jy~beam$^{-1}$.

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in {\bf length}. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). We immediately see from the high resolution VLBA imaging that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:model}. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total {\bf projected} mass within the Einstein radius (1.7~kpc for an assumed lens redshift of $z_l = 1$).  {\bf In this central region of an edge-on spiral galaxy, the baryonic mass can be a significant component of the total enclosed mass \citep[e.g.][]{swells2}.  In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the ExpDisc component as a good description of the baryonic mass of the lens galaxy.}  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models require {\bf the stronger constraints provided by deep imaging of the full extent of the gravitational arc in future high resolution radio observations.}

\begin{figure}
%\includegraphics[width=84mm]{1555_ao_merlin_overlay.eps}
\includegraphics[width=72mm]{fig2_twocontour_b.eps}
\caption{High-resolution $K^\prime$-band imaging of B1555+375 with contours from the MERLIN observations (red) of \citet{Marlow99} and the VLBA observations presented here (blue) overlaid.}
\label{fig:merlin}
\end{figure}

\begin{figure}
%\includegraphics[width=72mm]{gravlens_exp_try5_plot.eps}
\includegraphics[width=72mm]{gravlens_exp_try5_plot-2.eps}
%\includegraphics[width=72mm]{fig2_twocontour_b.eps}
\caption{Observed radio (blue pluses) and model-predicted (red open circles) image positions of B1555+375. The lens-plane critical curves are shown with the blue solid line and the source-plane caustics are shown with the red dashed line. The position of the source is at ($-0.1953$, $-0.1494$), which is marked by the red filled circle. The centroid positions of the two mass components of the lens (an SIE and an exponential disc) are marked by the black filled triangles.}
\label{fig:model}
\end{figure}

@@

\textbf{ 
Here, we find evidence for an edge-on disc in the flux-ratio anomaly
gravitational lens B1555+375 that can also account for the required
perturbation in the mass model.  Moreover, near-infrared AO imaging of
other lenses in the SHARP sample shows that B1555+375 is not the only
system having a significant disc component detection in the lensing
galaxy; these gravitational lenses will be investigated in future
work, but the images alone suggest that other systems may have similar
links between flux-ratio anomalies and disc components. Such
high-order deviations from a smooth mass distribution have already
been suggested previously \citep{evans03, congdon05}, but this is the
first time that an observed disc has been explicitly modelled to
account for a flux-ratio anomaly. However, it is important to realise
that by including observationally-motivated modifications to a
`smooth' mass distribution, such as luminous dwarf companions or
edge-on discs, in the mass models, we are still limited in our
interpretation if we can not determine how much of the anomalies are
due to these components. For our current analysis of B1555+375, we
have shown that the entire anomaly could be caused by the edge-on
disc.  However, it may be possible to independently constrain the mass
of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would test
whether the disc actually makes a sufficient contribution to the mass
model \citep[e.g. as in the case of B1933+503;][]{suyu12}. Also, in
the blue contours of Fig.~\ref{fig:merlin}, B1555+375 shows evidence
for an extended $\sim$ 100 mas gravitational arc at radio wavelengths
that crosses the edge-on disc. Further high resolution imaging with
very long baseline interferometry will provide the observational
constraints required to fully test whether the disc is sufficient to
cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies in some systems. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account 
\citep[e.g.][]{Xu15}. A number of systems with significant disc components are
already known.  These include, for example, the sample of 16 edge-on disks from SDSS found in the SWELLS survey \citep{swells1}.  This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.}
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.

@@

\section*{Acknowledgments}
We thank the referee for comments that improved the paper.  We are grateful to Dandan Xu for useful discussions. CDF and DJL acknowledge support from NSF-AST-0909119.  LVEK is supported in part through an NWO-VICI career grant (project number 639.043.308). The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc. Based on observations made with the NASA/ESA Hubble Space Telescope, obtained from the data archive at the Space Telescope Science Institute. STScI is operated by the Association of Universities for Research in Astronomy, Inc. under NASA contract NAS 5-26555. Some of the data presented herein were obtained at the W. M. Keck Observatory, which is operated as a scientific partnership among the California Institute of Technology, the University of California and the National Aeronautics and Space Administration. The Observatory was made possible by the generous financial support of the W. M. Keck Foundation. The authors wish to recognize and acknowledge the very significant cultural role and reverence that the summit of Mauna Kea has always had within the indigenous Hawaiian community.  We are most fortunate to have the opportunity to conduct observations from this mountain.

>>>>

A common feature of dark-matter only simulations of structure formation is the presence of thousands of subhaloes that are associated with larger mass haloes \citep[e.g.][]{Springel08}. However, observations of the Local Group show many fewer satellite galaxies than are predicted by the simulations, even when taking into account the latest discoveries of new dwarf galaxies \citep{DES15,Kop15} and completeness corrections arising from the limited sky coverage imposed by the Galactic Plane.  
This is the well-known {\it missing satellite problem} \citep{Klypin1999,Moore1999,S07}. In order to understand this discrepancy, it is imperative to build up a large sample of satellites/subhaloes in galaxies outside the Local Group. However, this process is challenging due to the 
%optical/infrared 
faintness of the satellite galaxies.  In fact, lower-mass satellites may not be able to retain the gas needed for ongoing star formation \citep[e.g.][]{P11}, rendering them effectively dark.

@@

B1555+375 was observed as part of SHARP using the NIRC2 camera on the W. M. Keck-II Telescope on 2012 May 16 (PI: Fassnacht).  The AO system was used, with the corrections derived from the laser guide star and a $R=14.4$-mag tip-tilt star located $\sim45$~arcsec from the lens.  The narrow camera mode was used, giving a field of view of $\sim10$~arcsec on a side and a pixel scale of 10~mas. Six dithered 300~s exposures were obtained in the $K^{\prime}$ filter.  The data were reduced with the standard SHARP pipeline, which is a python-based package that is a refinement of the process described by \citet{Auger_EELS1}.  A cutout of the final reduced image is shown in Fig.~\ref{fig:multiband} and also in Fig.~\ref{fig:3panel}a, where the contours from 5-GHz Multi-Element Radio Link Interferometer Network (MERLIN) imaging by \citet{Marlow99} and 1.66-GHz VLBA imaging (see below) are overlaid.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, 
%as can be seen by comparing the radio contours to the $K^\prime$-band emission 
(Fig.~\ref{fig:3panel}b), and the faint, but clearly visible emission from {\bf the lensing galaxy that appears to be an edge-on disc}, with a position angle of $\sim 10$~deg and an {\bf axis ratio $q \sim 0.2$}.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission from these images is likely due to dust extinction and additional demagnification due to lensing.

@@

B1555+375 was observed with the 10 telescopes of the VLBA at a central observing frequency of 1.66 GHz on 2000 March 21 (BN0009; PI: Norbury). The data were recorded at 128 Mbits~s$^{-1}$ and then correlated to produce two spectral windows with 8 MHz bandwidth, 16 channels and both circular polarizations (RR and LL). The observations were phase referenced using J1544+398 every 3 min over the total observing time of 3 h. The data were reduced in the standard manner using the {\sc vlbautlis} analysis pipeline that is part of the Astronomical Image Processing Software (AIPS). Imaging was done with the {\sc clean} algorithm in AIPS using natural weighting of the visibilities, and restored using an elliptical beam of size $9.7 \times 6.9$~mas at a position angle of $-7.6^\circ$ east of north. The final map (also shown in Fig.~\ref{fig:3panel}a) has an rms of 78~$\mu$Jy~beam$^{-1}$.

We find that lensed images A and B have been resolved into a gravitational arc that is $\sim100$~mas in {\bf length}. Image C is still compact on mas-scales, whereas image D is only marginally detected ($\sim6~\sigma$). 
%We immediately see from the high resolution VLBA imaging 
The VLBA imaging shows that image B has a smaller angular size than image A, which is consistent with a perturbation in the mass model being responsible for the flux-ratio anomaly (recall that gravitational lensing conserves surface brightness), as opposed to free-free absorption or interstellar scattering.

@@

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:3panel}c. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. The model requires that the ExpDisc component makes up about 10 per cent of the total {\bf projected} mass within the Einstein radius (1.7~kpc for an assumed lens redshift of $z_l = 1$).  {\bf In this central region of an edge-on spiral galaxy, the baryonic mass can be a significant component of the total enclosed mass \citep[e.g.][]{swells2}.  In the NIR imaging there is no sign of a bulge; a bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the ExpDisc component as a good description of the baryonic mass of the lens galaxy.}  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models require {\bf the stronger constraints provided by deep imaging of the full extent of the gravitational arc in future high resolution radio observations.}

\begin{figure*}
\includegraphics[width=0.9\textwidth]{b1555_3panel.eps}
\caption{
{\bf (a)} High-resolution $K^\prime$-band imaging of B1555+375 with contours from the MERLIN observations (red) of \citet{Marlow99} and the VLBA observations presented here (blue) overlaid.
%
{\bf (b)} Test, for now
%
{\bf (c)} Observed radio (blue pluses) and model-predicted (red open
circles) image positions of B1555+375. The lens-plane critical curves
are shown with the blue solid line and the source-plane caustics are
shown with the red dotted line. The position of the source is at
($-0.1953$, $-0.1494$), which is marked by the red filled circle. The
centroid positions of the two mass components of the lens (a SIE and
an exponential disc) are marked by the black filled triangles.
}
\label{fig:3panel}
\end{figure*}

% \begin{figure}
% %\includegraphics[width=84mm]{1555_ao_merlin_overlay.eps}
% \includegraphics[width=72mm]{fig2_twocontour_b.eps}
% \caption{High-resolution $K^\prime$-band imaging of B1555+375 with contours from the MERLIN observations (red) of \citet{Marlow99} and the VLBA observations presented here (blue) overlaid.}
% \label{fig:merlin}
% \end{figure}
% 
% \begin{figure}
% %\includegraphics[width=72mm]{gravlens_exp_try5_plot.eps}
% \includegraphics[width=72mm]{gravlens_exp_try5_plot-2.eps}
% %\includegraphics[width=72mm]{fig2_twocontour_b.eps}
% \caption{Observed radio (blue pluses) and model-predicted (red open circles) image positions of B1555+375. The lens-plane critical curves are shown with the blue solid line and the source-plane caustics are shown with the red dashed line. The position of the source is at ($-0.1953$, $-0.1494$), which is marked by the red filled circle. The centroid positions of the two mass components of the lens (an SIE and an exponential disc) are marked by the black filled triangles.}
% \label{fig:model}
% \end{figure}

@@

\textbf{ 
Here, we find evidence for an edge-on disc in the flux-ratio anomaly
gravitational lens B1555+375 that can account for the required
perturbation in the mass model.  Moreover, near-infrared AO imaging of
other lenses in the SHARP sample shows that B1555+375 is not the only
system that has a significant disc component detection in the lensing
galaxy; these gravitational lenses will be investigated in future
work, but the images alone suggest that other systems may have similar
links between flux-ratio anomalies and disc components. Such
high-order deviations from a smooth mass distribution have already
been suggested previously \citep{evans03, congdon05}, but this is the
first time that an observed disc has been explicitly modelled to
account for a flux-ratio anomaly. However, it is important to realise
that by including observationally-motivated modifications to a
`smooth' mass distribution, such as luminous dwarf companions or
edge-on discs, in the mass models, we are still limited in our
interpretation if we can not determine how much of the anomalies are
due to these components. For our current analysis of B1555+375, we
have shown that the entire anomaly could be caused by the edge-on
disc.  However, it may be possible to independently constrain the mass
of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would test
whether the disc actually makes a sufficient contribution to the mass
model \citep[e.g. as in the case of B1933+503;][]{suyu12}. Also, in
the blue contours of Fig.~\ref{fig:3panel}a, B1555+375 shows evidence
for an extended $\sim$ 100 mas gravitational arc at radio wavelengths
that crosses the edge-on disc. Further high resolution imaging with
very long baseline interferometry will provide the observational
constraints required to fully test whether the disc is sufficient to
cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies in some systems. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) be taken into account 
\citep[e.g.][]{Xu15}. A number of lens systems with significant disc components are
already known, including, for example, the sample of 16 edge-on disks from SDSS found in the SWELLS survey \citep{swells1}.  This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines \citep[e.g.][]{N14}.}
  %Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems \citep[and the presence of a disc-like component may not resolve the anomaly anyway;][]{evans03}, indicating that these systems are still useful for probing the dark matter substructure mass function.

@@

\section*{Acknowledgments}
We thank the referee for comments that improved the paper.  We are grateful to Dandan Xu for useful discussions. CDF and DJL acknowledge support from NSF-AST-0909119.  LVEK is supported in part through an NWO-VICI career grant (project number 639.043.308). The NRAO is a facility of the NSF operated under cooperative agreement by Associated Universities, Inc. Based on observations made with the NASA/ESA Hubble Space Telescope, obtained from the data archive at the Space Telescope Science Institute. STScI is operated by the Association of Universities for Research in Astronomy, Inc. under NASA contract NAS 5-26555. Some of the data presented herein were obtained at the W. M. Keck Observatory, which is operated as a scientific partnership among the California Institute of Technology, the University of California and the National Aeronautics and Space Administration. The Observatory was made possible by the generous financial support of the W. M. Keck Foundation. The authors wish to recognize and acknowledge the very significant cultural role and reverence that the summit of Mauna Kea has always had within the indigenous Hawaiian community.  We are most fortunate to have the opportunity to conduct observations from this mountain.
